Analysis and elimination of bottlenecks in parallel algorithm for solving global optimization problems

### 2022-04-14 2022/RSD/paper.tex

The novelty of the parallel algorithm considered in the present paper consists in the parallelization of the whole computing process -- both at the global search stage and at the stage of the local refining of the solution found. This approach can be relevant for the problems whose objective functions are flat valleys with a weak multiextremality. In such problems, the phase of searching the rough approximation of the global solution and the phase of its local refining are comparable in the computation costs.
For example, in the work \cite{Gubaydullin2021} the global search stage only was parallelized; the local refining was made with a sequential algorithm.

@@

Fig. XX presents the averaged numbers of iterations, which Hooke-Jeeves and BFGS methods executed for minimization of Rosenbrock function with the dimensionalities from 2 to 30. Fig. XX presents corresponding speedups in iterations. The number of parallel trials $p$ varied from 1 to 16.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{RosenbrocNT.jpg}
  \caption{Number of trials executed on the Rosenbrock function}
  \label{fig:RosenbrocNT}  
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{RosenbrocSpeedup.jpg}
  \caption{Speedup on the Rosenbrock function}
  \label{fig:RosenbrocSpeedup}  
\end{center}
\end{figure}

@@

Table XX presents the averaged number of iterations, which GSA required to solve the problem. Fig. presents the numbers of trials executed by Hooke-Jeeves and L-BFGS-B local methods. Fig. XX presents the speedups achieved when solving the problem as a whole.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{GKLSNT.jpg}
  \caption{Number of trials executed on the GKLS function}
  \label{fig:GKLSNT}  
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{GKLSSpeedup.jpg}
  \caption{Speedup on the GKLS function}
  \label{fig:GKLSSpeedup}  
\end{center}
\end{figure}

@@

To solve the problem stated, the synchronous parallelization scheme (using MPI) was employed. The number of processes $p$ to compute the function values varied from 1 up to 15. Also, there was an additional managing process, which processed the accumulated search information. In the stopping criteria of the method the value $\varepsilon=0.01$ was used and the reliability parameter of the method was set as $r = 3$. After the work of GSA, a parallel local method was started to refine the result found. The precision of the local method was set to $10^{-4}$ and the limitation on the number of trials was $10^4$.

%Table XX presents the total time of solving four problems corresponding to four different temperature regimes.
%
%\begin{table}[ht]
	%\caption{Averaged time of solving the problem, sec.}
	%\label{tab:11}
	%\center
		%\begin{tabular}{|c|c|c|c|}
		%\hline
		%%\noalign{\smallskip}
		%Method & P = 1 & P = 3 &  P = 15 \\
		%\hline 
		%Hooke-Jeeves & 2386.4 & 1053.1 & 713.1  \\
		%\hline
		%L-BFGS-B & 2571.7  & 1096.4  & 720.3  \\
		%\hline
	%\end{tabular}
%\end{table}

@@

\begin{table}[ht]
	\caption{Time of solving the problem using a local method}
	\label{tab:12}
	\center
		\begin{tabular}{|c|c|c|c|}
		\hline
		%\noalign{\smallskip}
		Method & GSA & GSA-l & GSA-pl \\
		\hline 
		Hooke-Jeeves & 158.1 & 299.8  & 177.4 \\
		\hline
		L-BFGS-B & 158.1  & 252.5  & 165.6  \\
		\hline
	\end{tabular}
\end{table}

Concluding general results, one can say that the use of local methods at the final stage of search allows solving the optimization problems with higher precision. In many cases, it is enough to use sequential local methods. However, for the problems with complex gully structure of the objective function require parallelizing the whole search process including the local stage as well. In the present work, two local optimization algorithms with the parallelization variants for shared memory as well as for distributed one are described. As the experiments have shown, the methods for parallelizing the local methods (at the final stage of global search) proposed demonstrated good results in solving both the test optimization problems and the applied ones.

>>>>

The novelty of the parallel algorithm considered in the present paper consists in the parallelization of the whole computing process -- both at the global search stage and at the stage of the local refining of the solution found. This approach can be relevant for the problems whose objective functions are ravine-type with a weak multiextremality. In such problems, the phase of searching the rough approximation of the global solution and the phase of its local refining are comparable in the computation costs.
For example, in the work \cite{Gubaydullin2021} the global search stage only was parallelized; the local refining was made with a sequential algorithm.

@@

Fig.~\ref{fig:RosenbrocNT} presents the averaged numbers of iterations, which Hooke-Jeeves and BFGS methods executed for minimization of Rosenbrock function with the dimensionalities from 2 to 30. Fig.~\ref{fig:RosenbrocSpeedup} presents corresponding speedups in iterations. The number of parallel trials $p$ varied from 1 to 16.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{RosenbrocNT.pdf}
  \caption{Number of trials (in thousands) executed on the Rosenbrock function}
  \label{fig:RosenbrocNT}  
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{RosenbrocSpeedup.pdf}
  \caption{Speedup on the Rosenbrock function}
  \label{fig:RosenbrocSpeedup}  
\end{center}
\end{figure}

@@

Table~\ref{tab:5} presents the averaged number of iterations, which GSA required to solve the problem. Fig.~\ref{fig:GKLSNT} presents the numbers of trials executed by Hooke-Jeeves and L-BFGS-B local methods. Fig.~\ref{fig:GKLSSpeedup} presents the speedups achieved when solving the problem as a whole.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{GKLSNT.pdf}
  \caption{Number of trials executed on the GKLS function}
  \label{fig:GKLSNT}  
\end{center}
\end{figure}

\begin{figure}
\begin{center}
  \includegraphics[width=1.0\linewidth]{GKLSSpeedup.pdf}
  \caption{Speedup on the GKLS function}
  \label{fig:GKLSSpeedup}  
\end{center}
\end{figure}

@@

To solve the problem stated, the synchronous parallelization scheme (using MPI) was employed. The number of processes to compute the function values was $p=15$. Also, there was an additional managing process, which processed the accumulated search information. In the stopping criteria of the method the value $\varepsilon=0.01$ was used and the reliability parameter of the method was set as $r = 3$. After the work of GSA, a parallel local method was started to refine the result found. The precision of the local method was set to $10^{-4}$ and the limitation on the number of trials was $10^4$.

%Table XX presents the total time of solving four problems corresponding to four different temperature regimes.
%
%\begin{table}[ht]
	%\caption{Averaged time of solving the problem, sec.}
	%\label{tab:11}
	%\center
		%\begin{tabular}{|c|c|c|c|}
		%\hline
		%%\noalign{\smallskip}
		%Method & P = 1 & P = 3 &  P = 15 \\
		%\hline 
		%Hooke-Jeeves & 2386.4 & 1053.1 & 713.1  \\
		%\hline
		%L-BFGS-B & 2571.7  & 1096.4  & 720.3  \\
		%\hline
	%\end{tabular}
%\begin{table}[ht]
%	\caption{Averaged time of solving the problem, sec.}
%	\label{tab:11}
%	\center
%		\begin{tabular}{|c|c|c|c|}
%		\hline
%		%\noalign{\smallskip}
%		Method & P = 1 & P = 3 &  P = 15 \\
%		\hline 
%		Hooke-Jeeves & 2386.4 & 1053.1 & 713.1  \\
%		\hline
%		L-BFGS-B & 2571.7  & 1096.4  & 720.3  \\
%		\hline
%	\end{tabular}
%\end{table}

@@

\begin{table}[ht]
	\caption{Time of solving the applied problem, sec.}
	\label{tab:12}
	\center
		\begin{tabular}{|c|c|c|c|}
		\hline
		%\noalign{\smallskip}
		Method & GSA & GSA-l & GSA-pl \\
		\hline 
		Hooke-Jeeves & 158.1 & 299.8  & 177.4 \\
		\hline
		L-BFGS-B & 158.1  & 252.5  & 165.6  \\
		\hline
	\end{tabular}
\end{table}

%\begin{table}
	%\caption{Time of solving the problem}
	%\label{tab:last}
	%\center
	%\begin{tabular}{llllllll}
		%\hline\noalign{\smallskip}
		%$p$ & \multicolumn{3}{l}{ Total time of solving four problems } & & \multicolumn{3}{l}{One of the temperature regimes only} \\
		%\noalign{\smallskip} \cline{2-4} \cline{6-8} \noalign{\smallskip}
		 %& P = 1 & P = 3 &  P = 15 & & GSA & GSA-l & GSA-pl  \\
		%\noalign{\smallskip} \hline \noalign{\smallskip}
		%Hooke-Jeeves & 2386.4 & 1053.1 & 713.1 & & 158.1 & 299.8  & 177.4  \\
		%L-BFGS-B & 2571.7  & 1096.4  & 720.3 & & 158.1  & 252.5  & 165.6  \\
%
		%\noalign{\smallskip}\hline
	%\end{tabular}
%\end{table}

Concluding the results, one can say that the use of local methods at the final stage of search allows solving the optimization problems with higher precision. In many cases, it is enough to use sequential local methods. However, the problems with ravine-type objective function require parallelizing the whole search process including the local stage as well. In the present work, two local optimization algorithms with the parallelization variants for shared memory as well as for distributed one are described. As the experiments have shown, the methods for parallelizing the local methods (at the final stage of global search) proposed demonstrated good results in solving both the test optimization problems and the applied ones.
